Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2019-12-18 main.tex
Adding edits

The use of multiple antennas at the transmitter and receiver can
significantly enhance the performance of wireless systems. Moreover,
in these multiple-input multiple-output (MIMO) wireless systems, the
transmitter can allocate resources more efficiently if the channel
state information (CSI) is made available at the transmitter, both to
improve data rates and to reduce BER~\cite{love2008overview}. To this
end, it is necessary to efficiently feedback the CSI from the receiver 
to the transmitter. In wireless systems that use orthogonal
frequency division multiplexing (OFDM), the feedback needs to be
provided to the transmitter across all subcarriers, and this places a
large feedback demand. However, since the CSI varies gradually in both
time and frequency, effective techniques to reduce feedback overheads
exist, using frequency domain interpolation temporal prediction. In
particular, knowledge of the precoding matrices (that are typically
the orthonormal right singular vectors of the channel matrix) is
essential to achieve link capacity, as well as to parallelize the
channel. In this paper, we focus on the problem of tracking the
precoders of a MIMO-OFDM system across time and frequency for
efficient CSI feedback by tracking a parametrized version of the matrices.

The problem of effective quantization and feedback of precoders for
MIMO-OFDM systems has attracted significant attention in recent
years. Many approaches involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency. While these
methods have been shown to be very effective for precoder feedback,
they involve operations over higher dimensional manifolds. In
particular, optimization and interpolation is complicated, especially
for manifolds where geodesic paths for interpolation may be difficult
to obtain, especially for joint time-frequency predictive coding. An
alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices parameterized into
independent scalar parameters that correspond to Givens rotations and
Householder transformations of the unitary precoding matrices
\cite{4114278,4556174}. These approaches have been combined with
adaptive delta modulation effectively track precoding matrices both in
the time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately.

In this paper, we propose a method for exploiting the joint
time-frequency correlation of the precoder's scalar parameters on the
transmitter using CSI feedback. Typical approaches to predictive
quantization and interpolation do not exploit the joint time-frequency
correlation to enhance the precoder estimate at the
transmitter. However, exploiting the correlations jointly not only
reduces the feedback requirement, but also enhances the precoder
reconstruction accuracy significantly. Further, the independence of
the scalar parameters reduces the problem to one that involves
interpolation of separate scalar valued functions, as opposed to
operations over tangent spaces in
manifolds~\cite{Gupt1905:Predictive}. Simulations reveal that the
proposed method matches the performance of manifold based approaches
with a much lower implementation complexity and fewer number of bits
for quantization.

The rest of this paper is organized as follows: Section~\ref{section2}
describes the system model and the precoder feedback and interpolation
techniques. Section~\ref{section3} compares the BER and achievable
rates for our proposed methods with other approaches. Finally,
Section~\ref{section4} concludes with discussions on future work.

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, the channel for the $i$-th subcarrier at time instant $t$
can be modelled as a flat-fading MIMO channel. This flat-fading MIMO
channel can be modelled as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ \bf{\eta_{i,t}}
\end{equation}
where $\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$ is the channel
matrix, $\bx_{i,t} \in \mathbb{C}^{N_d}$ is the input data vector,
$\by_{i,t} \in \mathbb{C}^{N_R}$ is the output vector,
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ is the unitary
precoder, and $\eta$ is the complex additive white Gaussian noise
distributed as ${\mathcal{CN}}(0,I_{N_R})$. We consider the case where
$N_R < N_T$, and the data vector size $N_d = N_R$.  The ``thin'' SVD
(that keeps only those right and left singular vectors that correspond
to non-zero singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_R} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\\mathcal{CN}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier. However, the frequency coherence and
temporal correlations would reduce the overall feedback requirement,
as discussed in subsequent sections.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model for
for the COST207 channel power delay
profiles~\cite{molisch2006cost259,cost1989cost}. The BER achieved by
the proposed predictive quantization based precoder is compared
against the ideal precoder for 100 channel evolutions in each run and
averaged. Simulations are performed for normalised Doppler values of
$f_dT_s = 3.5\times 10^{-6}$ (corresponding to a velocity of 36 km/h
for $f_c = 1$ GHz) with 1024 subcarriers, and with
$f_dT_s = \times 1.05 \times 10^{-5}$ (velocity of 108 km/h) with 64
subcarriers. The sampling time period $T_s$ was $5\times10^{-8}$
s. For both the situations, we considered $N_T=4$ transmit antennas
and and $N_R=2$ receive antennas. Therefore, the total number of
parameters that determine the precoder is
$N_{R}(2N_{T} - N_{R}) = 12$, with five $\theta_l$s and and seven
$\phi_i$s. We assume that the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the fast and slow channel profiles
respectively (i.e. for the slower channels with 1024 subcarriers, the
subcarriers indices fed back would be $0, 33, 66, \ldots 1023$ and for
the faster channel, with 64 subcarriers, $0, 9, 18, \ldots 63$ would
be fed back).

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 2 bits for
initializing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times 2$ = 20 (we note that $\phi_0 = \phi_4 = 0$ due to the
non-uniqueness of the SVD). Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, either the $\theta_l$ or the $\phi_i$ parameters
are fed back alternately, while retaining the unsent parameters as
being to their previous values, as shown in
Fig.~\ref{fig:adpm-fig}. Here, $\Theta$ is the collection of all
$\theta_l$s and $\Phi$ is the collection of all $\phi_i$s. This brings
down the bit budget to an average of 5 bits(the length of $\Theta$)
and 7 bits (the length of $\Phi$) over alternate feedback instances,
thereby resulting in an average feedback budget of 5 bits.

>>>>

The use of multiple antennas at the transmitter and receiver (MIMO) can
significantly enhance the performance of wireless systems. 
A typical step to get this performance enhancements is estimation of channel
state information (CSI).
If transmitter also has the knowledge of CSI, it can allocate 
resources more efficiently, both to
improve data rates and to reduce BER~\cite{love2008overview}.

To this end, it is necessary to efficiently feedback the CSI from the receiver 
to the transmitter. In wireless systems that use orthogonal
frequency division multiplexing (OFDM), the feedback needs to be
provided to the transmitter across all subcarriers and in each OFDM frame, 
which places a large feedback demand. However, since the CSI varies gradually in both
time and frequency, effective techniques to reduce feedback overheads
exist, using frequency domain interpolation and temporal prediction [CITE?]. In
particular, knowledge of the precoding matrices (that are typically
the orthonormal right singular vectors of CSI matrix) is
essential to achieve the link capacity, as well as to parallelize the
channel. In this paper, we focus on the problem of tracking the
precoders of a MIMO-OFDM system across time and frequency for
efficient CSI feedback by tracking a parametrized version of the precoding matrices.

The problem of effective quantization and feedback of precoders for
MIMO-OFDM systems has attracted significant attention in recent
years. Many approaches involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency. 
An alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices parameterized into
independent scalar parameters that correspond to Givens rotations and
Householder transformations of the unitary precoding matrices
\cite{4114278,4556174} [?].

While the manifold based methods have been shown to be very effective for precoder feedback,
they involve operations over higher dimensional manifolds. In
particular, optimization and interpolation is complicated, especially
for manifolds where geodesic paths for interpolation may be difficult
to obtain, especially (don't repeat especially again) for joint time-frequency predictive coding.
In addition to being analytically more tractable, the scalar parametrization methods have also been
implemented in 802.11ac standards \cite{porat2013improved}. Past work in utilizing
scalar parametrization for precoder feedback have utilized
adaptive delta modulation effectively track precoding matrices both in
the time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately. In this paper, we propose a method for exploiting the joint
time-frequency correlation of the precoder's scalar parameters on the
transmitter using CSI feedback. 
Typical approaches to predictive
quantization and interpolation do not exploit the joint time-frequency
correlation to enhance the precoder estimate at the
transmitter. However, exploiting the correlations jointly not only
reduces the feedback requirement, but also enhances the precoder
reconstruction accuracy significantly.

Most of the papers utilizing manifolds \cite{Gupt1905:Predictive,pitaval2013codebooks,6891198}  and scalar parametrization \cite{4114278,4556174} compare the performance to other manifold/scalar parametrization methods, but don't do the cross comparison to highlight the performance benefits of the chosen method.
In addition to a novel joint time-frequency feedback scheme, we compare also the results of our
algorithm with an equivalent manifold based scheme, and illustrate the performance
benefits entailed by utilizing the scalar parametrization method. 
%  Further, the independence of
% the scalar parameters reduces the problem to one that involves
% interpolation of separate scalar valued functions, as opposed to
% operations over tangent spaces in
% manifolds~\cite{Gupt1905:Predictive}. 
Simulations reveal that the
proposed method matches the performance of manifold based approaches
with a much lower implementation complexity and fewer number of bits
for quantization.

% The rest of this paper is organized as follows: Section~\ref{section2}
% describes the system model and the precoder feedback and interpolation
% techniques. Section~\ref{section3} compares the BER and achievable
% rates for our proposed methods with other approaches. Finally,
% Section~\ref{section4} concludes with discussions on future work.

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, the channel for the $i$-th subcarrier at time instant $t$
can be modelled as a flat-fading MIMO channel. This flat-fading MIMO
channel can be written as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ \bf{\eta_{i,t}}
\end{equation}
where $\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$ is the channel
matrix, $\bx_{i,t} \in \mathbb{C}^{N_d}$ is the input data vector,
$\by_{i,t} \in \mathbb{C}^{N_R}$ is the output vector,
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ is the unitary
precoder, and $\eta$ is the complex additive white Gaussian noise
distributed as ${\mathcal{CN}}(0,I_{N_R})$. We consider the case where
$N_R < N_T$, and the data vector size $N_d = N_R$.  The ``thin'' SVD
(that keeps only those right and left singular vectors that correspond
to non-zero singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_R} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\mathcal{CN}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier. However, the frequency coherence and
temporal correlations would reduce the overall feedback requirement,
as discussed in subsequent sections.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model for
for the COST207 channel power delay
profiles~\cite{molisch2006cost259,cost1989cost}. The BER achieved by
the proposed predictive quantization based precoder is compared
against the ideal precoder for 100 channel evolutions in each run and
averaged. Simulations are performed for normalised Doppler values of
$f_dT_s = 3.41\times 10^{-6}$ (corresponding to a velocity of 37.2 km/h
for $f_c = 1$ GHz) with 1024 subcarriers, and with
$f_dT_s = 1.56 \times 10^{-5}$ (velocity of 172.8 km/h) with 64
subcarriers. The sampling time period $T_s$ was $5\times10^{-8}$
s. For both the situations, we considered $N_T=4$ transmit antennas
and and $N_R=2$ receive antennas. Therefore, the total number of
parameters that determine the precoder is
$N_{R}(2N_{T} - N_{R}) = 12$, with five $\theta_l$s and and seven
$\phi_i$s. We assume that the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the slow and fast channel profiles
respectively (i.e. for the slower channels with 1024 subcarriers, the
subcarriers indices fed back would be $0, 33, 66, \ldots 1023$ and for
the faster channel, with 64 subcarriers, $0, 9, 18, \ldots 63$ would
be fed back).

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 2 bits for
initializing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times 2$ = 20 (we note that $\phi_0 = \phi_4 = 0$ due to the
non-uniqueness of the SVD). Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, either the $\theta_l$ or the $\phi_i$ parameters
are fed back alternately, while retaining the unsent parameters as
being to their previous values, as shown in
Fig.~\ref{fig:adpm-fig}. Here, $\Theta$ is the collection of all
$\theta_l$s and $\Phi$ is the collection of all $\phi_i$s. This brings
down the bit budget to an average of 5 bits(the length of $\Theta$)
and 7 bits (the length of $\Phi$) over alternate feedback instances,
thereby resulting in an average feedback budget of 6 bits.
